### 2023-02-28 calculus.tex
Brief section on hazelnut action calculus

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
indicate a type mismatch error on the other.

@@

Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
such ill-typed ones meaningless. This is insufficient for editor services and language servers,
which must provide users feedback on where and why a program is ill-typed. Moreover, they must
continue providing editor services even when an error is encountered. Indeed, similarly to
error-recovering parsers, real-world type checkers must recover from type errors. Traditionally,
this has been performed in an \emph{ad hoc} manner [citations?], with different languages and tools
adopting varying approaches in the localization, reportage, and recovery from errors.

@@

The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
type error localization in bidirectional systems, a judgmental framework that provides a unified
theory for the design of typing semantics that supports ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a minimal gradually
typed lambda calculus. All rules and theorems may be found in the supplementary appendix, alongside
a complete mechanization in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the system might be
extended to more complex typing features, such as System F-style polymorphism.

@@

\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
system in which type information is recursively propagated through terms and allows for local type
inference.

@@

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory.

% TODO include holes
\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV} \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of the extended gradually typed lambda calculus}
  \label{fig:calculus-syntax}
\end{figure}

@@

\section{Marking Calculus in Structured Editing}
\label{sec:calculus-structured-editing}

@@

\subsection{Implementation in Hazel}

@@

>>>>

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
``blame'' the type mismatch error on the other.

@@

Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
such ill-typed ones meaningless. This is insufficient for editor services and language servers,
which must provide users feedback on where and why a program is ill-typed. Moreover, they must
continue providing semantic services even when an error is encountered. Indeed, similarly to
error-recovering parsers, real-world type checkers must robustly recover from type errors.
Traditionally, this has been performed in an \emph{ad hoc} manner [citations?], with different
languages and tools adopting varying approaches in the localization, reportage, and recovery of
errors.

@@

The key contribution of this section is the \emph{marking calculus}, a formal calculus for type
error localization in bidirectional systems. Namely, the calculus provides a judgmental framework
with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus, extending it with binary product types in \cref{sec:calculus-products}. All rules
and theorems may be found in the supplementary appendix, alongside a complete mechanization in the
Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we then
consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

% \emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
% type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
% system in which type information is recursively propagated through terms and allows for local type
% inference.

@@

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory. $\EEHole$ denotes an \emph{expression
hole}, which are used to model syntactically incomplete programs---but are not critical within the
calculus.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \EEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended gradually typed lambda calculus.}
  \label{fig:calculus-syntax}
\end{figure}

@@

\section{The Marking Calculus in Structured Editing}
\label{sec:calculus-structured-editing}

@@

\subsection{Implementation in Hazel}
\label{sec:calculus-hazel}

@@

The Hazelnut action calculus introduced by \citet{HazelnutPOPL} give a structure editing scheme that
ensures programs are statically meaningful.
% TODO say something more about Hazelnut, especially in relation to error holes

Hazelnut, however, is unable to cope with actions that require non-local hole fixes. % TODO example of this
Such actions remain undefined, i.e. unperformable.

By applying the marking calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
from the typing semantics. Under this new system, the action calculus is concerned only with the
manipulation of syntax, and the marking procedure yields statically meaningful terms with error
holes inserted at the appropriate positions. Furthermore, marking solves the problem of non-local
hole fixes; actions may be performed anywhere (as long as they are of the correct sort), and marking
totality ensures that they give meaningful terms.
% TODO this is not strictly true, only for 'wrapping' actions
